---
title: 'Time Series: A First Course with Bootstrap Starter'
output: pdf_document
toc: true
toc-depth: 3
---

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = "Data")
```

\pagebreak

# Lesson 2: Takes and tools 

## Takes 

add law of cosines 

- triangle 
- vectors 
- form in the book: $cos(\theta)$ expresses autocorrelation 

elements from Pradel 

- les 3 perp 

gaussian random V comp avec pradel

- cumul de (0,1) donne une N(0,1) vect

- multiplication par une racine de mat de var-covar (equivalent du $\sigma$)

pour simuler $\mathcal{N}(\mu,\Sigma)$

## Questions

- def matrice var covar before props: vv'

- relation avec ps: v'v

- pb of bivariate normal (here ?)

- cholesky decomp = square root, not unique (à un scalire pres ou "pire")

- student moy/ var

## Links to JD+ 

algos, doc and related training

in my training 

in tooklkit or other functions ? 

- distribs ?

in GUI display ? 

## Overview in book 


### Concepts 


### R skills

- simulate bivariate, n variate normal ?



# Lesson 2-1: Random Vectors

- A time series sample is a finite stretch of realizations, i.e., a vector.
- A vector of random variables is called a random vector: $\underline{X} = {[X_1, \ldots, X_n]}^{\prime}$.

## Mean and Covariance

- The mean of $\underline{X}$ is a vector, each of whose components is the expectation
of the corresponding random variable: ${\mathbf E} [ X_i]$.
- The covariance matrix of $\underline{X}$ has entries given by the covariance between
the corresponding components of the random vector: $\mbox{Cov} [X_j, X_k]$.
- The covariance matrix $\mbox{Cov} [\underline{X}]$ of a random vector is non-negative definite
and symmetric. Its eigenvalues are real and non-negative.

## Affine Transforms

- $\underline{Y} = A \underline{X} + \underline{b}$ is an affine transform of $\underline{X}$.
- If ${\mathbf E} \underline{X} = \underline{\mu}$ and $\mbox{Cov} [\underline{X}] = \Sigma$,
then ${\mathbf E} \underline{Y} = A \underline{\mu} + \underline{b}$ and
$\mbox{Cov} [\underline{Y}] = A \Sigma A^{\prime}$.

## Covariance Decomposition

- We can decompose a symmetric matrix $\Sigma$ as
\[
 \Sigma = P \Lambda P^{\prime}
\]
for an orthogonal matrix $P$ (i.e., $P^{\prime} = P^{-1}$), and where $\Lambda$ is 
a diagonal matrix consisting of the real eigenvalues of $\Sigma$.
- A symmetric non-negative definite matrix $\Sigma$ can be decomposed as $\Sigma = B B^{\prime}$,
and $B$ is called a square root (it is not unique). One such square root is the *Cholesky* factor.
- If $\underline{Z}$ has i.i.d. components with mean zero and variance one, then 
$\underline{X} = B \underline{Z} + \underline{\mu}$ is a random vector with mean $\underline{\mu}$
and covariance matrix $B B^{\prime}$. 

### Simulation Example

- Simulate a bivariate random vector with mean $[1,2]$ and covariance matrix
\[
\Sigma = \left[ \begin{array}{cc} 2 & 1 \\ 1 & 4 \end{array} \right].
\]

```{r}
Sigma <- rbind(c(2,1),c(1,4))
mu <- c(1,2)
B <- t(chol(Sigma)) # t to get lower triangular
z <- matrix(rnorm(2*100),nrow=2) # 2 vecteurs de 100 obs
x <- B %*% z + mu
print(colMeans(t(x)))
print(var(t(x)))
```

## Gaussian Random Vectors

- A random vector $\underline{Y}$ is Gaussian with mean $\underline{\mu}$ and 
non-singular covariance matrix $\Sigma$ if its joint pdf is
\[
 p_{\underline{Y}} (\underline{y}) = {(2 \pi)}^{-n/2} {(\det \Sigma)}^{-1/2}
  \exp \{ - {( \underline{y} - \underline{\mu} )}^{\prime} \Sigma^{-1}
   {( \underline{y} - \underline{\mu} )} /2   \}.
\]
- Denoted by writing $\underline{Y} \sim \mathcal{N} (\underline{\mu}, \Sigma)$.
- An affine transformation of a Gaussian vector is still Gaussian. In particular, 
sub-vectors are Gaussian.
- We can decorrelate a Gaussian random vector: $\underline{X} = B^{-1} \underline{Y}$
has $\mbox{Cov} [\underline{X}] = B^{-1} \Sigma B^{-1 \prime} = 1_n$, the identity
matrix.
- The quadratic form 
\[
{(\underline{Y} - \underline{\mu})}^{\prime} \Sigma^{-1} {(\underline{Y} - \underline{\mu})}
\]
 has a $\chi^2$ distribution on $n$ degrees of freedom.
 
 # Lesson 2-2: Stochastic Processes

- A collection of random variables indexed by time is called a *stochastic process*, denoted as $\{ X_t \}$. The curly brackets let us know $\{ X_t \}$ is the process, whereas $X_t$ is a single random variable (at time $t$).
- Usually time is $t \in {\mathbf Z}$, the integers.
- There are also continuous-time stochastic processes (another subject).

## Realization is Sample Path

- A random variable $X_t$ has realization $x_t$.
- Example: $X_t \sim \mathcal{N}(0,1)$ has realization `r set.seed(123); rnorm(1)`.
- Put this together for all $t \in {\mathbf Z}$, and the realization is called the *sample path*.

### Example: Heavy-tailed Sample Path

```{r}
set.seed(777)
n <- 100
z <- rt(n+1,df=4)		# heavy-tailed input
theta <- .8
x <- z[-1] + theta*z[-(n+1)]
plot(ts(x),xlab="Time",ylab="")
```

- As usual, we connect the dots when graphing the sample path.
- Here is another realization, or sample path, of the same stochastic process.

```{r}
set.seed(888)
n <- 100
z <- rt(n+1,df=4)		# heavy-tailed input
theta <- .8
x <- z[-1] + theta*z[-(n+1)]
plot(ts(x),xlab="Time",ylab="")
```

## Common Examples of Stochastic Processes

### Example 2.2.8. Process A: i.i.d.

- An i.i.d. process, where each $X_t$ has the same distribution and is independent of the rest.

```{r}
set.seed(111)
n <- 100
x <- runif(n)
plot(ts(x),xlab="Time",ylab="")
```

### Example 2.2.9. Process B: Cosine

- Suppose $X_t = A \cos (\vartheta t + \Phi)$, where $\vartheta$ is given, and $A$ and $\Phi$ are independent random variables.

```{r}
n <- 100
set.seed(222)
A <- rnorm(1)
set.seed(223)
phi <- 2*pi*runif(1)
lambda <- pi/6
set.seed(224)
x <- A*cos(seq(1,n)*lambda + phi)
plot(ts(x),xlab="Time",ylab="")
```

### Example 2.2.12. Process E: Random Walk

- Suppose $X_t$ is current location on a straight line, and we step forward or backward at time $t+1$. Let the step size be given by random variable $Z_{t+1}$, independent of where we are. Then our new location is
\[ 
 X_{t+1} = X_t + Z_{t+1}.
\]
 This is called a *random walk*.
- We can initialize with $X_0 = 0$, for example.
 
```{r}
set.seed(333)
n <- 100
z <- rnorm(n)
x <- rep(0,n)
x0 <- 0
x[1] <- x0 + z[1]
for(t in 2:n) { x[t] <- x[t-1] + z[t] }
plot(ts(x),xlab="Time",ylab="")
```


# Lesson 2-3: Stationarity

- We want to generalize the concept of *identical distribution* to a stochastic process.

## Marginal Distributions

- First marginals are just the $X_t$ random variables' distributions.
- Second marginals are joint distributions for all pairs $(X_t,X_s)$.
- Third marginals are joint distributions for all triplets, etc.

### Same First Marginals

- Saying $\{ X_t \}$ has same first marginal is same as saying they are identically distributed.
- Sometimes called *First Order Stationary*.
- In particular, all means are the same: ${\mathbf E} [X_t] = {\mathbf E} [X_s]$ for all $t,s$.

### Second Marginals Under Shift

- Suppose all pairs have the same distribution when shifted: 
\[
  (X_1, X_2) \sim (X_2, X_3) \sim (X_3, X_4) \ldots
\]
- Then second marginal distribution only depends on lag $h$, i.e., distribution of $(X_t, X_{t-h})$ does not depend on $t$.
- Sometimes called *Second Order Stationary*.
- Then the product mean (the covariance) depends only on lag:
\[
 {\mathbf E} [ X_t \, X_{t-h}]. 
\]
It does not depend on $t$.  
 
## Example: Visualizing Stationarity

- We generate 100 simulations of a Gaussian AR(1), and generate a scatterplot of $(X_1, X_2)$ 
- We repeat with $(X_3, X_4)$
 
```{r}
x1 <- NULL
x2 <- NULL
x3 <- NULL
x4 <- NULL

for(i in 1:100) {

z <- rnorm(10)
x <- rep(0,10)
phi <- .9
x0 <- rnorm(1)/sqrt(1-phi^2)
x[1] <- phi*x0 + z[1]
for(t in 2:10) { x[t] <- phi*x[t-1] + z[t] }

x1 <- c(x1,x[1])
x2 <- c(x2,x[2])
x3 <- c(x3,x[3])
x4 <- c(x4,x[4])
}

plot(x2,x1,xlab="X Past",ylab="X Present")

plot(x4,x3,xlab="X Past",ylab="X Present")
```

# Lesson 2-4: Autocovariance

- Now we study the autocovariance function.

## Strict and Weak Stationarity

- Strict stationarity: all marginals (of all orders) are time shift invariant.
- Weak stationarity: the time series has finite variance, constant mean $\mu$, and covariance only depends on lag $h$:
\[
  \gamma (h) = \mbox{Cov} [X_t, X_{t-h}] = {\mathbf E} [ X_t \, X_{t-h}] - \mu^2.
\]
This function is called the **autocovariance**.
- So the variance is $\gamma (0)$.
- The **autocorrelation** is
\[
 \rho (h) = \frac{ \gamma (h) }{ \gamma (0)}.
 \]
- Weak stationarity is sometimes called *covariance stationarity*. 
 
## Example: Autocorrelation of an AR(1)

- We plot $\rho (h)$ versus $h$ (on x-axis).
 
```{r}
phi <- .8
rho <- phi^seq(0,20)
plot(ts(rho,start=0),xlab="Lag",ylab="Rho",type ="h")
```

## White Noise

- A key example is a *white noise* stochastic process.
- This is any weakly stationary process $\{ Z_t \}$ with mean zero such that $\gamma (h) = 0$ for $h \neq 0$.
- Written compactly as $Z_t \sim \mbox{WN} (0, \sigma^2)$, where $\sigma^2 = \gamma (0)$ is the variance, and the mean is $\mu = 0$.

## Covariance Matrix of Sample Vector

- The time series variables corresponding to a sample are $X_1, \ldots, X_n$, which can be put into a random vector $\underline{X}$.
- The covariance matrix of $\underline{X}$ is denoted by $\Gamma_n$ when the stochastic process is weakly (or strictly) stationary. The entry in row $j$ and column $k$ is
\[
  \Gamma_n (j,k) = \mbox{Cov} [ X_j, X_k] = \gamma (k-j).
\]
 This only depends on the difference between row and column index! Such a matrix is constant along diagonals, and is called *Toeplitz*.
 
```{r}
rho <- .8
gamma <- rho^seq(0,5)/(1-rho^2)
gamma_mat <- toeplitz(gamma)
gamma_mat
```

## Properties of Autocovariance

1. $\gamma (0) \geq 0$
2. $\gamma (h) = \gamma (-h)$
3. $|\gamma (h)| \leq \gamma (0)$.
4. $\gamma (h)$ is a non-negative definite sequence.

This last property means that $\Gamma_n$ is a non-negative definite matrix for all $n$. (Recall from multivariate analysis: covariance matrices are non-negative definite, and are positive definite if all eigenvalues are positive.)


# Lesson 2-5: Autoregression and Moving Average

- Examples of weakly stationary stochastic process.

## Example 2.5.1. AR(1) Process

- Let $Z_t \sim \mbox{i.i.d.} (0,\sigma^2)$ and $\{ X_t \}$ defined via
\[
 X_t = \phi \, X_{t-1} + Z_t
\]
for $t \geq 1$, where $|\phi| < 1$.
- This is a recursion, called an *order 1 autoregression*, or AR(1).
- How to define $X_0$? If $X_0 \sim (0, \sigma^2/(1-\phi^2) )$, then $\{ X_t \}$ is weakly stationary and
\[
 \gamma (h) = \sigma^2 \frac{ \phi^{|h|}}{1 - \phi^2}.
\]
 Formula only makes sense when $|\phi| < 1$. This is a *stationarity condition*.
 
```{r}
n <- 100
set.seed(123)
z <- rnorm(n)
x <- rep(0,n)
phi <- .9
x0 <- rnorm(1)/sqrt(1-phi^2)
x[1] <- phi*x0 + z[1]
for(t in 2:n) { x[t] <- phi*x[t-1] + z[t] }
plot(ts(x),xlab="Time",ylab="")
```

## Example 2.5.5. MA(1) Process

- Let $Z_t \sim \mbox{i.i.d.} (0,\sigma^2)$ and $\{ X_t \}$ defined via
\[
 X_t =  Z_t + \theta \, Z_{t-1}
\]
for $t \geq 1$, where $\theta$ is any real number.
- This process is called an *order 1 moving average*, or MA(1). 
- It is weakly stationary, with
\[
 \gamma (h) = \begin{cases} (1 + \theta^2) \sigma^2 \quad \mbox{if} \; h = 0 \\
  \theta \sigma^2 \quad \mbox{if} \; h = \pm 1 \\
  0 \quad \mbox{if} \; |h| > 1.
  \end{cases}
\]

```{r}
set.seed(777)
n <- 100
z <- rnorm(n+1)		# Gaussian input
theta <- .8
x <- z[-1] + theta*z[-(n+1)]
plot(ts(x),xlab="Time",ylab="")
```

## Example 2.5.6. MA(2) Process

- Let $Z_t \sim \mbox{i.i.d.} (0,\sigma^2)$ and $\{ X_t \}$ defined via
\[
 X_t =  Z_t + \theta_1 \, Z_{t-1} + \theta_2 \, Z_{t-2}
\]
for $t \geq 2$, where $\theta_1, \theta_2$ are any real numbers.
- This process is called an *order 2 moving average*, or MA(2). 
- It is weakly stationary, with
\[
 \gamma (h) = \begin{cases} (1 + \theta_1^2 + \theta_2^2) \sigma^2 \quad \mbox{if} \; h = 0 \\
  (\theta_1 + \theta_1 \theta_2) \sigma^2 \quad \mbox{if} \; h = \pm 1 \\
  \theta_2 \sigma^2 \quad \mbox{if} \; h = \pm 2\\
  0 \quad \mbox{if} \; |h| > 2.
  \end{cases}
\]

```{r}
set.seed(555)
n <- 100
z <- rnorm(n+2)		# Gaussian input
theta1 <- .9
theta2 <- .2
x <- z[-c(1,2)] + theta1*z[-c(1,n+2)] + theta2*z[-c(n+1,n+2)]
plot(ts(x),xlab="Time",ylab="")
```


# Lesson 2-6: White Noise Processes

- *White noise* is a fundamental building block for time series models.
- Any $\{ X_t \}$ i.i.d. with mean zero and variance $\sigma^2$ is a $\mbox{WN}(0,\sigma^2)$.
- Here we provide three examples of white noise. 


## Example 2.6.1. Dependent White Noise
  
- Consider $X_t = Z_t \cdot Z_{t-1}$, where $Z_t$ is i.i.d. $N(0,1)$.  
- Then $X_t \sim \mbox{WN} (0,1)$, but $\{ X_t \}$ is not i.i.d.

```{r}
n <- 101
z <- rnorm(n)
x <- z[-n]*z[-1]
plot(ts(x),xlab="Time",ylab="")
#acf(x)
```

## Example 2.6.2. Non-identically Distributed White Noise

- Let $\{ Y_t \}$ and $\{ Z_t \}$ be independent of each other. Set $Z_t$ i.i.d. $N(0,1)$ and $Y_t$ i.i.d. uniform on $(-\sqrt{3},\sqrt{3})$. Let 
\[
 X_t = \begin{cases} Z_t \quad t \; \mbox{even} \\ Y_t \quad t \; \mbox{odd} \end{cases}
\]
- Then $X_t \sim \mbox{WN} (0,1)$, although the process is not stationary (since the marginal distribution depends on $t$).

```{r}
n <- 100
z <- rnorm(n)
y <- runif(n,-3^(1/2),3^(1/2))
x <- matrix(t(cbind(y,z)),ncol=1)
plot(ts(x),xlab="Time",ylab="")
#acf(x)
```


## Example 2.6.3. ARCH Process

- Model of Engel (1982), a Nobel laureate.
- Set $Z_t$ i.i.d. $N(0,1)$, and $X_t = Z_t \, \sqrt{\alpha + \beta X_{t-1}^2}$.
- Then $X_t \sim \mbox{WN} (0,\alpha/(1-\beta))$.

```{r}
n <- 101
z <- rnorm(n)
alpha <- .2
beta <- .3
x <- 0
for(t in 2:n) { x <- c(x,z[t]*sqrt(alpha + beta*x[t-1]^2)) }
plot(ts(x),xlab="Time",ylab="")
#acf(x)
```



